Towards Practical First-Order Model Counting

### 2025-01-22 doc/paper/paper.tex
Significantly expand the introductory part of Section 3

\todo[inline,caption={}]{Add a discussion of the innovation aspect and technical
  parts. Motivate each part of \Cranetwo{} (e.g., how \FindBaseCases addresses
  key issues in \textsc{Crane}).}

@@

    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (findbasecases);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (propagate);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{The outline of using \Cranetwo{} to compute the model count of a
    formula $\phi$. First, the formula is compiled into a set of equations,
    which are then used to create a C++ program. This program can be executed
    with different command line arguments to calculate the model count of $\phi$
    for different domain sizes. To accomplish this, the \CompileWithBaseCases
    procedure makes use of the FOKC algorithm of \textsc{Crane}, algebraic
    simplification techniques (denoted as \Simplify), and two other auxiliary
    procedures. }\label{fig:overview}
\end{figure*}

\Cref{fig:overview} provides an overview of \Cranetwo{}'s workflow.
\Cref{sec:completing} describes the main algorithm for completing the
definitions of recursive functions with a sufficient set of base cases.
\Cref{sec:identifying,sec:propagating} describe subsidiary algorithms for
constructing a set of base cases and their corresponding logical formulas.
\Cref{sec:smoothing} explains the post-processing techniques for ensuring
accurate model counting. Additionally, \cref{sec:cpp} explains the process of
compiling equations into C++ code, greatly expanding upon the range of formulas
that could previously be handled by similar
approaches~\cite{DBLP:conf/kr/KazemiP16}.

@@

\emph{Smoothing} modifies a circuit to reintroduce eliminated atoms, ensuring
the correct model
count~\cite{darwiche2001tractable,DBLP:conf/ijcai/BroeckTMDR11}. In this
section, we describe a similar process performed on
lines~\ref{line:presmoothing}--\ref{line:smoothing2} of \cref{alg:propagate}.
Line~\ref{line:presmoothing} checks if smoothing is necessary, and
lines~\ref{line:smoothing1} and~\ref{line:smoothing2} execute it. If the
condition on \autoref{line:presmoothing} is not satisfied, the clause is not
smoothed but omitted.

>>>>

@@

    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (findbasecases);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (propagate);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{The outline of using \Cranetwo{} to compute the model count of a
    formula $\phi$. First, the formula is compiled into a set of equations,
    which are then used to create a C++ program. This program can be executed
    with different command line arguments to calculate the model count of $\phi$
    for different domain sizes. To accomplish this, the \CompileWithBaseCases
    procedure uses \textsc{Crane}, algebraic simplification techniques (denoted
    as \Simplify), and two other auxiliary procedures. }\label{fig:overview}
\end{figure*}

\Cref{fig:overview} provides an overview of \Cranetwo{}'s workflow. Below we
briefly describe and motivate each procedure before going into more detail in
the corresponding subsection.

\CompileWithBaseCases (see \cref{sec:completing}), the core procedure of
\Cranetwo{}, is responsible for completing the function definitions produced by
\textsc{Crane} with the necessary base cases. To do so, it may recursively call
itself (and \textsc{Crane}) on other formulas. We prove that the number of such
recursive calls is upper bounded by the number of domains in the formula.

\cref{sec:completing} also describes the \Simplify procedure for algebraic
simplification. It is crucial for simplifying, e.g., a sum of $n$ terms, only
two of which are non-zero. More generally, the equations returned by
\textsc{Crane} often benefit from easy-to-detect algebraic simplifications such
as $0 \cdot \text{anything} = 0$ and $\text{anything}^{0} = 1$.

\FindBaseCases (described in \cref{sec:identifying}) inspects a set of equations
to identify a sufficient set of base cases for a given set of equations. We
prove that the returned set of base cases is sufficient, and the evaluation of
the resulting function definitions will never get stuck in an infinite loop.

\Cref{sec:propagating} introduces the \Propagate procedure that takes a formula
$\phi$, a domain $\Delta$, and $n \in \mathbb{N}_{0}$. It returns $\phi$
transformed with the assumption that $|\Delta| = n$, introducing $n$ new
constants and removing all variables quantified over $\Delta$. For example, when
computing a base case such as $f(0, y)$, \Propagate will significantly simplify
$\phi$ with the assumption that the domain associated with the first parameter
of $f$ (i.e., $\mathcal{D}(f, 1)$) is empty. When run on this simplified
formula, \CompileWithBaseCases will return the equations for the base case
$f(0, y)$.

\Cref{sec:smoothing} describes a new kind of \emph{smoothing} used to ensure
that \Propagate preserves the correct model count. Smoothing is a well-known
technique in knowledge compilation algorithms for propositional model
counting~\cite{darwiche2001tractable}. Although it has been applied to FOMC
before~\cite{DBLP:conf/ijcai/BroeckTMDR11}, our setting requires a novel
approach.

\CompileWithBaseCases, together with the other procedures outlined above, return
a set of equations that fully cover the base cases of all recursive functions.
While these equations can be interesting and valuable in their own right, the
users of FOMC algorithms typically expect a numerical answer. Thus,
\cref{sec:cpp} describes how these equations are compiled into a C++ program
that can be executed with different command-line arguments to compute the model
count for different combinations of domain sizes.

@@

Smoothing modifies a circuit to reintroduce eliminated atoms, ensuring the
correct model count~\cite{darwiche2001tractable,DBLP:conf/ijcai/BroeckTMDR11}.
In this section, we describe a similar process performed on
lines~\ref{line:presmoothing}--\ref{line:smoothing2} of \cref{alg:propagate}.
Line~\ref{line:presmoothing} checks if smoothing is necessary, and
lines~\ref{line:smoothing1} and~\ref{line:smoothing2} execute it. If the
condition on \autoref{line:presmoothing} is not satisfied, the clause is not
smoothed but omitted.
